{RandomRationalPoints} package for \emph{Macaulay2}

### 2021-04-02 FastLinAlg/M2/ChangesMade_ResponseToReviewers.txt

a) Section 2.1 (Page 2),

b) 2.2 Page 3;

c) Sentence before Example 2.5 added on Page 4

d) Page 5 Remark 2.2.1 added following reviewer's suggestion to discuss a bit about the probability of succesful output.

@@

d) Page 5 Section 5 dimViaBezout added with example 3.1 and paragraph following it.

MAIN CHANGES:

-- We implement the dimViaBezout to compute the dimension probabilistically. This intersects V(I) with successively higher dimensional random linear spaces until there is an intersection.  For example, if $V(I)$ intersect a random line has a point, then we expect that $V(I)$ contains a hypersurface.  If there was no intersection, this function tries a 2-dimensional linear space, and so on.  This speeds up many computations.

@@

Optional Inputs:

DimensionIntersectionAttempts => ZZ : the number of linear spaces to try before moving to the next dimension
MinimumFieldSize => ZZ: if the ambient field is smaller than this value it will automatically be replaced with an extension. (The user may set the  MinimumFieldSize to ensure that the field being worked over is big enough.  For instance, there are relatively few linear spaces over a field of characteristic 2, and this can cause incorrect results to be provided.)

-- The reviewer suggested the use of dimDegViaBezout where the use of saturation is needed. We avoid this issue (which may be potentially decelerating) totally by the above successive intersections.

-- The MultiplicationTable strategy has also beeen added. (This MultiplicationTable strategy essentially is still a LinearIntersection but it handles the point issue differently. The previous strategy was using decompose to break up the ideal.)

-- Following the reviewer's suggestion, we have added Remark 2.2.1 with references that all point to a good probability of finding a rational point by intersecting with a line in the case of hypersurface.

### 2021-04-19 FastLinAlg/M2/ChangesMade_ResponseToReviewers.txt

b) 2.2 Page 3;

c) Sentence before Example 2.5 added on Page 4

d) Page 5 Remark 2.2.1 added following reviewer's suggestion to discuss a bit about the probability of succesful output.

e) Page 5 Section 5 dimViaBezout added with example 3.1 and paragraph following it.

@@

MAIN CHANGES:

-- We have implemented the dimViaBezout to compute the dimension probabilistically. This intersects V(I) with successively higher dimensional random linear spaces until there is an intersection.  For example, if $V(I)$ intersect a random line has a point, then we expect that $V(I)$ contains a hypersurface.  If there was no intersection, this function tries a 2-dimensional linear space, and so on.  This speeds up many computations.

    Some comments on implementation.  The referee suggested using a binary search pattern, however we found that in practice, it is much faster to start with a 1-dimensional linear space, then a 2-dimensional linear space, and so on, until we find an intersection.  This is because in our examples, checking whether an intersection is empty is much faster when the intersection is indeed empty.

    Furthermore, our implementation is in the affine case and not the projective setting (it did not seem to impact performance, and now it works on a wider range of examples).

@@

-- Following the reviewer's suggestion, we have added Remark 2.2.1 with references that all point to a good probability of finding a rational point by intersecting with a line in the case of hypersurface.

Changes Made to the Source Code:

We have added the MultiplicationStrategy verbatim as provided by the reviewer but have changed the computation of the dimension to be done via dimViaBezout instead of dimDegViaBezout. (This MultiplicationTable strategy essentially is still a LinearIntersection but it handles the point issue differently. The previous strategy was using decompose to break up the ideal.)

There are some aspects of this code we do not totally understand.  In particular, we do not understand what is occurring in the lines:

m0:=contract(transpose matrix entries b2,matrix entries((xx_0*b1)%Js));
and what follows.

Additionally, we have had quite a few examples, especially when running something iteratively, where the saturation:
Js:=saturateInGenericCoordinates J;
(which only saturates in the last coordinate)
will saturate away the points, and thus causes this function to crash.

However, doing a saturation with respect to every variable (even separately) seems to make this function quite a bit slower.  We are tempted to saturate with respect to a (random) variable, then use a try command to see if that yields a good result, and if it crashed try saturating with respect to a different (random) variable.  However, we

@@

New documentation and examples have been provided.

All the strategies now make use of the dimViaBezout to compute the dimension of the given ideal, if the codimension is not already known. In case, that it is known, it should be specified along with the call of randomPoints.

ProjectionAttempts has been set to 0 in the Default Strategy now for reasons discussed in

### 2021-04-19 FastLinAlg/Paper/RandomRationalPoints.tex
Fixes to the paper and to the comments

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage}
% \usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\hypersetup{
  bookmarks,
  bookmarksdepth=3,
  bookmarksopen,
  bookmarksnumbered,
  pdfstartview=FitH,
  colorlinks,backref,hyperindex,
  linkcolor=Sepia,
  anchorcolor=BurntOrange,
  citecolor=MidnightBlue,
  citecolor=OliveGreen,
  filecolor=BlueViolet,
  menucolor=Yellow,
  urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
% \usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
%\usepackage{enumitem}
%\SetLabelAlign{parleft}{\parbox[t]\textwidth{#1\smallskip}}

@@

    \begin{itemize}
    \item {\tt Default} performs a sequence of the different strategies below, aimed at finding a point quickly.  It begins with less general projections and linear intersections, and gradually ramps up the randomness.  
    \item {\tt BruteForce} simply tries random points and sees if they are on the variety.

    \item {\tt GenericProjection} projects to a hypersurface, via 
        {\tt projectionToHypersurface} and then uses the {\tt BruteForce} strategy.

@@

    \item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersection.

    \item {\tt MultiplicationTable} works for homogeneous ideals only. It cuts down with a linear space to a zero-dimensional projective scheme and computes how the last two variables act on the quotient ring truncated  
    at the regularity. In case of an absolutely irreducible $I$ over the field, we will find with  high probability a point, 
    since the determinant will have a linear factor in at least 50\% of the cases.
    \end{itemize}

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Even for some simple examples, it could not provide any rational points if the number of trials is not enough. Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly. Providing {\tt codimension} speeds up computations significantly as computing codimension using the default command {\tt codim} creates a bottleneck. If codimension needs to be computed, we  circumvent this issue using a probabilistic code to compute it (see \Cref{helper}). A quick comment, while in many examples generic projection works well, there are some examples where it performs quite slowly.  Setting {\tt ProjectionAttempts=>0} in the {\tt Default} strategy is a way to avoid this situation primarily.

@@

@@

\vspace{1em}
\end{document}

>>>>

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage}
% \usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\hypersetup{
  bookmarks,
  bookmarksdepth=3,
  bookmarksopen,
  bookmarksnumbered,
  pdfstartview=FitH,
  colorlinks,backref,hyperindex,
  linkcolor=Sepia,
  anchorcolor=BurntOrange,
  citecolor=MidnightBlue,
  citecolor=OliveGreen,
  filecolor=BlueViolet,
  menucolor=Yellow,
  urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
% \usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\newtheorem{remark}{Remark}[subsection]

\newcommand{\bA}{\mathbb{A}}
\newcommand{\bF}{\mathbb{F}}
%\usepackage{enumitem}
%\SetLabelAlign{parleft}{\parbox[t]\textwidth{#1\smallskip}}

@@

    \begin{itemize}
    \item {\tt Default} performs a sequence of the different strategies below, aimed at finding a point quickly.  It begins begins with brute force and moves to linear intersections with particularly simple linear forms, and gradually ramps up the randomness.  If {\tt ProjectionAttempts} is set to a value bigger than zero (it is set to zero by default), then the strategy alternates between generic projects and hybrid projection / intersections.  For additional discussion, see \cref{rem.ProjectingToAHypersurfaceFirst}.  
    \item {\tt BruteForce} simply tries random points and sees if they are on the variety.

    \item {\tt GenericProjection} projects to a hypersurface, via 
        {\tt projectionToHypersurface} and then uses the {\tt BruteForce} strategy.  Finally, it pulls back back the points that were found to the original variety (via a linear intersection).  See \cref{rem.ProjectingToAHypersurfaceFirst}.

@@

    \item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersectio with that hypersurface.  Finally, it pulls back back the points that were found to the original variety (via a linear intersection).   See \cref{rem.ProjectingToAHypersurfaceFirst}.  
    \item {\tt MultiplicationTable} works for homogeneous ideals only. It cuts down with a linear space to a zero-dimensional projective scheme and computes how the last two variables act on the quotient ring truncated at the regularity.  %In case of an absolutely irreducible $I$ over the field, we will find with  high probability a point, since the determinant will have a linear factor in at least 50\% of the cases.
    \end{itemize}

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Even for some simple examples, it could not provide any rational points if the number of trials is not enough. Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly. Providing {\tt codimension} speeds up computations significantly as computing codimension using the default command {\tt codim} creates a bottleneck. If codimension needs to be computed, we  circumvent this issue using a probabilistic code to compute it (see \Cref{helper}). A quick comment, while in many examples generic projection works well, there are some examples where it is many orders of magnitude slower.  By default, generic projection is turned off in the default strategy, that is we set {\tt ProjectionAttempts=>0}.  Increasing this number will include it.

@@

\begin{remark}[Finding a point]
	In the case of an absolutely irreducible hypersurface in $\bA^n_{\bF_q}$ (defined by $f$ say), there is significant discussion in the literature estimating lower bounds of number of rational points (see for instance, \cite{RationalKPointsSource,lang1954number,ghorpade2002number,cafure2006improved}) all of which point to the fact that there is ``good probability" of finding a rational point in this case when we intersect with a random line. Heuristically, we can make the following rough estimation.  We expect that each equation $f = \lambda$ for $\lambda \in \bF_q$ has approximately the same number of solutions.  Since each point on $\bF_q^n$ solves exactly one of these equations, we expect that $f = 0$ has approximately $q^{n-1}$ solutions, or in other words, our hypersurface has $q^{n-1}$-points.  
       Now, a random line $L$ has $q$ points. We want to find the probability that one of these points is  rational for $V(f)$.  We would expect that if these points are randomly distributed, then the probability that our line contains one of those points $1-(1-{1 \over q})^q$ which tends to $1-e^{-1} \approx 0.63$ for $q$ large.   Alternately, one can use the proof of \cite[Proposition 2.12]{bothmer2005quick} for a more precise statement.  For each point of $L$, we see that the probability that  the chosen point does not lie in the intersection, $L\cap V(f)$, is $1-\frac{1}{q}$. We then exhaust this search over all the points on $L$ to get the probability that there is indeed a successful intersection is $1-(1-\frac{1}{q})^{q}$. As $q$ gets larger, this value tends to $1-e^{-1}\approx 0.63$.

       Of course, there are certainly examples schemes over $\bF_q$ with no rational points at all, even for curves in $\bA^2$.
\end{remark}

\begin{remark}[Projecting to a hypersurface first]
    \label{rem.ProjectingToAHypersurfaceFirst}
    Suppose $X \subseteq \bA^n$ is an algebraic set.  In a number of existing algorithms, one first does a generic (or even not very generic) projection $h : \bA^n \to \bA^m$ and so that $h(X)$ is a hypersurface (at least set theoretically).  Then we find a point $x \in h(X)$ (say as above), and compute $h^{-1}(\{ x \})$ which is a linear space in $\bA^n$ that typically intersects $X$ in a rational point.   For example, this is done in {\tt randomKRationalPoint} in core Macaulay2.  Note that projecting to a hypersurface \emph{still is performing an intersection with a linear space}, but it just tries to choose the linear space more intelligently.

    However, while in a number of examples, doing this generic projection first can speed up computations (as opposed to simply intersecting with a linear space, since we have identified a linear space initially), there are numerous cases where computing this hypersurface $h(X)$ can be extremely slow.  This particularly appears in cases when one is computing successive minors to identify the locus where some variety is nonsingular.  Thus, by default, we disable the functionality to project to a hypersurface in the default strategy.

    On the other hand, instead of using a truly random linear space to intersect with, in the default strategy, we normally choose a linear space whose defining equations are particularly non-complicated (ie, with many linear or binomial terms).  Such simple linear spaces are the ones considered in {\tt randomKRationalPoint} for instance.  In practice, this seems give approximately the same performance as projecting to a hypersurface, without the chance of the code getting hung up on the generic projection.
\end{remark}

@@

\end{document}
